Thermodynamic response functions in a cell fluid model

### 2024-09-13 paper_text.tex
add ukr abstract
update reference to Fig1

Figure~\ref{fig1} shows the isotherms for the pressure $P^*$ versus density $\rho^*$ (see Fig.~\ref{fig1}a), and versus the effective chemical potential $M$ (see Fig.~\ref{fig1}b).
\begin{figure}[h!]
	\centering 
	\includegraphics[width=0.446\textwidth]{f1a1.pdf} 
	\includegraphics[width=0.45\textwidth]{f1b1.pdf} 
	\vskip-3mm\caption{Isotherms of the reduced pressure $P^*$ as a function (a) of the density $\rho^*$, and (b) of the effective chemical potential $M$ at $T \geq T_c$ represented by black lines. Thick grey lines on both figures correspond to isotherms of pressure at $T < T_c$ based on the results taken from~\cite{KozlovskiiDobush2020}.
	}\label{fig1}
\end{figure}

@@

First, we provide an explicit example of differentiating the equation of state, namely~\eqref{eq:eosPTn_reduced}, by deriving the expression for the thermal pressure coefficient $\beta^*_V$ from~\eqref{eq:beta_star_n}.
For the quantity $\bar{M}$ from Eq.~\eqref{eq:M_nT} one has
\begin{equation}
	\left( \frac{\partial \bar{M}}{\partial \tau} \right)_{\rho^*} =
	\frac{k_{\rm B}T_c \rho_n}{W(0)} \frac{1 - \omega_0}{1 + \tau\omega_0},
\end{equation}
where
\begin{equation*}
	\omega_0 = \frac{\chi_0 + A_{\gamma}}{B - 1 + \chi_0}.
\end{equation*}
For $d$ from~\eqref{def:D0} one has
\begin{equation}
	\frac{\partial d}{\partial \tau} = -\frac{k_{\rm B}T_c}{W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0}.
\end{equation}
While calculating both derivatives, we take into account
\begin{equation}
	\frac{\partial}{\partial \tau} \left[\beta W(0)\right] = -\frac{\beta_c \Phi^{(r)}(0)}{(1+\tau)^2} (B - 1 - A_{\gamma}).
\end{equation}
The derivative of the quantity $E_\rho$ from~\eqref{eq:E_nu} with respect to temperature $\tau$ is
\begin{eqnarray}
	\frac{\partial E_\rho}{\partial \tau} & = &  -\frac{1}{2}\left(\bar{M} - g_1 + \frac{g_3}{g_4} d + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\frac{\partial \beta W(0)}{\partial \tau} 
	\nonumber\\
	&&-\beta W(0)\left(\bar{M} - g_1 + \frac{g_3}{g_4} d + \frac{g_3^3}{6g_4^2} \right)
	\nonumber\\
	&& \times \left(\frac{\partial \bar{M}}{\partial \tau} + \frac{g_3}{g_4}\frac{\partial d}{\partial \tau}\right)
	\nonumber\\
	&& -\frac{g_3}{g_4}\frac{\partial \bar{M}}{\partial \tau} 
	- \frac{g_3^2}{2g_4^2}\frac{\partial d}{\partial \tau}.
\end{eqnarray}
Collecting these all formulas together, we can explicitly calculate $\beta^*_V$ from~\eqref{eq:beta_star_n}
\begin{eqnarray}
	\beta^*_V & = & E_\rho (\rho^*,T) + \bar{M} \rho_{n} + \frac{d}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4
	\nonumber\\
	&& + (1+\tau) \left(\frac{\partial E_{\rho}}{\partial \tau} + \frac{\partial \bar{M}}{\partial \tau}\rho_{n} + \frac{\rho_n^2}{2}\frac{\partial d}{\partial \tau}\right).
\end{eqnarray}
Substituting the expressions for derivatives and grouping similar terms, one finally arrives at
\begin{eqnarray}
	\beta^*_V & = & E_\rho (\rho^*,T) + \bar{M} \rho_{n} + \frac{d}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4 + k_{\rm B} T_c (1+\tau)
	\nonumber\\
	&& \times \left\{ \frac{\beta^2}{2}  \Phi^{(r)}(0) [B - 1 - A_\gamma]	\left(\bar{M} - g_1+ \frac{g_3}{g_4} d + \frac{1}{6} \frac{g_3^3}{g_4^2} \right)^2 \right . \nonumber \\
	&& \left . + \frac{1}{W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0} \left[ \frac{g_3^2}{2 g_4^2} + \frac{\rho_n^2}{2} - \frac{g_3}{g_4} \rho_n  \right . \right . \nonumber \\
	&& \left . \left . + \beta W(0) \left( \bar{M} - g_1+ \frac{g_3}{g_4} d + \frac{1}{6} \frac{g_3^3}{g_4^2} \right) \left( \frac{g_3}{g_4} - \rho_n \right)    \right] \right\}.
\end{eqnarray}

Second, we derive the explicit expression for the isothermal compressibility $\kappa^*_T$ using~\eqref{eq:kappa_star_m}, taking the density $\rho^*$ from \eqref{eq:density}.
For the effective chemical potential $M$ from \eqref{chem_pot} one has
\begin{equation}
\left(\frac{\partial M}{\partial \mu} \right)_T = \frac{1}{W(0)}.
\end{equation}
For the quantity $\bar \rho_0$ from \eqref{eq:ro_MT} one has
\begin{equation}
\left(\frac{\partial \bar \rho_0}{\partial \mu} \right)_T = \frac{1}{ g_4 W(0) \sqrt{Q_t}} \left[ \bar \rho_0 - 2 \left( - \frac{3M}{g_4} + \sqrt{Q_t} \right)^{\frac{1}{3}} \right].
\end{equation}
Using these two formulas we obtain the isothermal compressibility in explicit form  
\begin{eqnarray}
	\kappa_T^* & = & \frac{\epsilon}{\rho^{*2} W(0)} \nonumber \\
	&& \times \left\{  \frac{1}{\beta W(0) g_4 \sqrt{Q_t}} \left[ \bar \rho_0 - 2 \left( - \frac{3M}{g_4} + \sqrt{Q_t} \right)^{\frac{1}{3}} \right] - 1 \right\}.
\end{eqnarray}

>>>>

Figure~\ref{fig1} shows the isotherms for the pressure $P^*$ as a function of density $\rho^*$ (see Fig.~\ref{fig1}a), and the effective chemical potential $M$ (see Fig.~\ref{fig1}b).
\begin{figure}[h!]
	\centering 
	\includegraphics[width=0.446\textwidth]{f1a1.pdf} 
	\includegraphics[width=0.45\textwidth]{f1b1.pdf} 
	\vskip-3mm\caption{Isotherms of the reduced pressure $P^*$ as a function (a) of the density $\rho^*$, and (b) of the effective chemical potential $M$ at $T \geq T_c$ represented by black lines. Thick grey lines on both figures correspond to isotherms of pressure at $T < T_c$ based on the results taken from~\cite{KozlovskiiDobush2020}.
	}\label{fig1}
\end{figure}

@@

First, we provide an explicit example of differentiating the equation of state, namely~\eqref{eq:eosPTn_reduced}, by deriving the expression for the thermal pressure coefficient $\beta^*_V$ from~\eqref{eq:beta_star_n}.
For the quantity $\bar{M}$ from Eq.~\eqref{eq:M_nT} one has
\begin{equation}
	\left( \frac{\partial \bar{M}}{\partial \tau} \right)_{\rho^*} =
	\frac{k_{\rm B}T_c \rho_n}{W(0)} \frac{1 - \omega_0}{1 + \tau\omega_0},
\end{equation}
where
\begin{equation*}
	\omega_0 = \frac{\chi_0 + A_{\gamma}}{B - 1 + \chi_0}.
\end{equation*}
For $d$ from~\eqref{def:D0} one has
\begin{equation}
	\frac{\partial d}{\partial \tau} = -\frac{k_{\rm B}T_c}{W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0}.
\end{equation}
While calculating both derivatives, we take into account
\begin{equation}
	\frac{\partial}{\partial \tau} \left[\beta W(0)\right] = -\frac{\beta_c \Phi^{(r)}(0)}{(1+\tau)^2} (B - 1 - A_{\gamma}).
\end{equation}
The derivative of the quantity $E_\rho$ from~\eqref{eq:E_nu} with respect to temperature $\tau$ is
\begin{eqnarray}
	\frac{\partial E_\rho}{\partial \tau} & = &  -\frac{1}{2}\left(\bar{M} - g_1 + \frac{g_3}{g_4} d + \frac{g_3^3}{6g_4^2} \right)^{2} 
	\frac{\partial \beta W(0)}{\partial \tau} 
	\nonumber\\
	&&-\beta W(0)\left(\bar{M} - g_1 + \frac{g_3}{g_4} d + \frac{g_3^3}{6g_4^2} \right)
	\nonumber\\
	&& \times \left(\frac{\partial \bar{M}}{\partial \tau} + \frac{g_3}{g_4}\frac{\partial d}{\partial \tau}\right)
	\nonumber\\
	&& -\frac{g_3}{g_4}\frac{\partial \bar{M}}{\partial \tau} 
	- \frac{g_3^2}{2g_4^2}\frac{\partial d}{\partial \tau}.
\end{eqnarray}
Collecting these all formulas together, we can explicitly calculate $\beta^*_V$ from~\eqref{eq:beta_star_n}
\begin{eqnarray}
	\beta^*_V & = & E_\rho (\rho^*,T) + \bar{M} \rho_{n} + \frac{d}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4
	\nonumber\\
	&& + (1+\tau) \left(\frac{\partial E_{\rho}}{\partial \tau} + \frac{\partial \bar{M}}{\partial \tau}\rho_{n} + \frac{\rho_n^2}{2}\frac{\partial d}{\partial \tau}\right).
\end{eqnarray}
Substituting the expressions for derivatives and grouping similar terms, one finally arrives at
\begin{eqnarray}
	\beta^*_V & = & E_\rho (\rho^*,T) + \bar{M} \rho_{n} + \frac{d}{2} \rho_{n}^2 - \frac{a_4}{24} \rho_{n}^4
	\nonumber\\
	& + & \left\{ \frac{\beta}{2}  \Phi^{(r)}(0) [B - 1 - A_\gamma]	G_M^2 \right . \nonumber \\
	& + & \left. \frac{1}{\beta W(0)} \frac{1 - \omega_0}{1 + \tau \omega_0} \left[ \frac{g_3^2}{2 g_4^2} + \frac{\rho_n^2}{2} - \frac{g_3}{g_4} \rho_n  \right . \right . \nonumber \\
	& + & \left. \left. \beta W(0) G_M \left( \frac{g_3}{g_4} - \rho_n \right)    \right] \right\},
\end{eqnarray}
where we introduced notation
\begin{equation*}
	G_M = \bar{M} - g_1+ \frac{g_3}{g_4} d + \frac{1}{6} \frac{g_3^3}{g_4^2}.
\end{equation*}

Second, we derive the explicit expression for the isothermal compressibility $\kappa^*_T$ using~\eqref{eq:kappa_star_m}, taking the density $\rho^*$ from \eqref{eq:density}.
For the effective chemical potential $M$ from \eqref{chem_pot} one has
\begin{equation}
\left(\frac{\partial M}{\partial \mu} \right)_T = \frac{1}{W(0)}.
\end{equation}
For the quantity $\bar \rho_0$ from \eqref{eq:ro_MT} one has
\begin{equation}
\left(\frac{\partial \bar \rho_0}{\partial \mu} \right)_T = \frac{1}{ g_4 W(0) \sqrt{Q_t}} \left[ \bar \rho_0 - 2 \left( - \frac{3M}{g_4} + \sqrt{Q_t} \right)^{\frac{1}{3}} \right].
\end{equation}
Using these two formulas we obtain the isothermal compressibility in explicit form  
\begin{eqnarray}
	\kappa_T^* & = & \frac{\epsilon}{\rho^{*2} W(0)} \bigg\{ -1
	\\
	& + & \left. \frac{Q_t^{-1/2}}{\beta W(0) g_4} \left[ \bar \rho_0 - 2 \left( - \frac{3M}{g_4} + \sqrt{Q_t} \right)^{\frac{1}{3}} \right]\right\}.
	\nonumber
\end{eqnarray}
